---
title: "Appendix of tutorial analyzing cyclic models"
author: " "
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### ---------------------------------------------------------------------------

### Preliminary Steps

There are three functions used in this appendix that should be activated by running the code. These functions can be downloaded from Github: https://github.com/PeterVerboon/Cyclic-models.
 
 function cycpar():computes the cyclic parameters from the linear model parameters <br>
 function fitCyclic(): fits a linear model to obtain cyclic parameters and provides plots <br>
 function fitCyclicMLA(): fits a MLA model to obtain cyclic parameters and provides plots  <br>


These functions should be helpful for applying a cyclic model to your data. The functions have methods for printing the results and plotting the fitted cyclic patterns.

&nbsp;

As an example we generate cyclic data and construct Figure 1 with these data. First we need to install and load the package ggplot2 necessary for plotting. 

```{r tutorial-0, eval=TRUE}

require(ggplot2)              

a0 <- 1;
b1 <- 2;
b2 <- 10;
P <- 24;

T <- seq(0, 24, by =0.1);
y <- a0 + b1*cos(2*(pi/P)*(T - b2)) 

g <- ggplot() + geom_line(aes(x=T, y=y)) ;
g <- g + labs(x = "Time points", y = "Dependent variable ");
g <- g + theme(axis.text = element_text(size = 10, colour="black"));
g <- g + scale_x_continuous(breaks = seq(0,24,2)) ;
g <- g + geom_line(aes(x=b2, y=seq(1,3,0.1)), lty="dashed") ;
g <- g + geom_line(aes(x=seq(0,24,0.1), y=a0)) ;
g <- g + annotate("text", x = 1, y = 1.2, label = "b0");
g <- g + annotate("text", x = 10.8, y = 2, label = "b1");
g <- g + annotate("text", x = 10, y = 0.8, label = "b2");
g
```

The "smoking" data are in the object `smokingdat.rda`. This is a dataframe, which contains a selection of variables from a larger data file. Check the names in the dataframe and then make put the data in dat1, the first data set used in the analysis.

```{r tutorial-1, eval=TRUE}
require(cyclicESM)
data("smokedat")
names(smokedat)

dat1 <- smokedat
nall <- length(unique(dat1$subjnr))

```

We count the number of records per subject and remove subjects with less than 50 records. The resulting data.frame is called dat3.

```{r tutorial-3, eval=TRUE}
dat1$count <- 1
dat2 <- aggregate(dat1[,c("count")],by=list(dat1$subjnr), FUN=sum, na.rm=T);
dat2$subjnr <- dat2$Group.1
dat1$count <- NULL;
dat2$count <- dat2[,2]

dat3 <- merge(dat1,dat2[,c("subjnr","count")], by.x = "subjnr")
rm(dat2)

dat3 <- subset(dat3, dat3$count >= 50)
nfew <- nall - length(unique(dat3$subjnr))

```

We also compute variability of intention per subject and remove subjects with very small variation in intention. The result is also in dat3.

```{r tutorial-4, eval=TRUE}

dat2 <- aggregate(dat3$intention,by=list(dat3$subjnr), FUN=sd, na.rm=F)
dat2$subjnr <- dat2$Group.1
dat2$Group.1 <- NULL
dat2 <- merge(dat3, dat2,  by.x = "subjnr")
dat3 <- subset(dat2, dat2$V1 > .10)  
rm(dat2)
nsmall <- (nall - nfew) - length(unique(dat3$subjnr))

```


```{r tutorial-printn, eval=TRUE, echo=FALSE}

cat(" Number of subjects in data set:              ", nall," \n",
    "Number of subjects with less than 50 records: ", nfew ," \n" ,
    "Number of subjects with SD smaller than .10:  ", nsmall," \n" ,
    "Number of subjects used in analysis:         ",length(unique(dat3$subjnr)))

```



The next step is to construct a dataset that is aggregated over subjects, which is called dat4. This dataframe will be used in subsequent analyses.

```{r tutorial-5, eval=TRUE}
dat4 <- aggregate(dat3[,c("positiveAffect","stress","intention")],
                  by=list(dat3$beepnr,dat3$daynr), FUN=mean, na.rm=F);    
dat4$daynr <- dat4$Group.2
dat4$beepnr <- dat4$Group.1
dat4$Group.1 <- NULL
dat4$Group.2 <- NULL
```

&nbsp;


###    ANALYSES  

#### Step 1

The raw data for three subjects and average are plotted. This makes Figure 2 from the tutorial.

```{r tutorial-6, eval=TRUE}
pdat <- dat4                                         # averaged over all subjects
pdat <- subset(dat3, dat3$subjnr == 2)               # ppn 2, 15, 18 selected for Figure 1
                                                      
npoints <- dim(pdat)[1]
x <- c(1:npoints)
pdat$day <- as.factor(pdat$daynr)

g <- ggplot(pdat,aes(x=x, y=pdat$intention, colour=pdat$day)) + geom_point() +
     scale_x_discrete(name ="Time points (beeps within days)",  labels=pdat$beepnr, limits=c(1:npoints)) +
     theme(axis.text = element_text(size = 6, colour="black"),legend.position="none") 
g

```

We have to repeat step 1 for the aggregated data in respectively dat4, and for subjects 2, 15 and 18.

 &nbsp;  
 
 
  
#### Step 2

Now we analyze the cyclic model and plot the result. This is Figure 3 from the tutorial.

```{r tutorial-7, eval=TRUE} 

model_a <- fitCyclic(pdat,yvar = "intention",  xvar1 ="beepnr", xvar2 = "daynr", 
                     ymin = -2.5, ymax = 1.5, step=.30)

print(model_a)
plot(model_a)
```
  
 &nbsp;  
 

#### Step 3

Fit an extra term in the model with day as covariate in aggregated data. Show the parameters and the fit of the model. 

```{r tutorial-8, eval=TRUE} 

model_b <- fitCyclic(pdat,  yvar = "positiveAffect", xvar1 ="beepnr",xvar2 = "daynr", 
                     cov = "daynr", ymin=-0.5, ymax=0.5)

print(model_b)
plot(model_b)
```

&nbsp; 


#### Step 4

Apply the cylic model with daily period for the variables stress and positive affect for subject 15 to obtain a plot of the raw data with the fitted cycles and a plot of the aggregated data. Together with intention this makes Figure 4.

```{r tutorial-9, eval=TRUE} 

pdat <- subset(dat3, dat3$subjnr == 15)   

model_c <- fitCyclic(pdat,yvar = "stress", xvar1 ="beepnr", xvar2 = "daynr", 
                     ymin = -1.0, ymax = 0.5, step= 0.25)

print(model_c)
plot(model_c)
```

&nbsp; 



#### Step 5

Apply, only for subject 15, the cylic model with a weekly period for stress and positive affect, instead of a daily period. This is shown in Figure 5.

```{r tutorial-10, eval=TRUE} 
pdat <- subset(dat3, dat3$subjnr == 15)   

model_d <- fitCyclic(pdat, yvar = "stress", xvar1 ="daynr", xvar2 = "daynr", 
                     ymin = -2.0, ymax = 1.0, step= 0.25)

print(model_d)
plot(model_d, type = "means")

```

&nbsp; 



#### Step 6

Analyze the data with the cyclic model using  MLA and model comparison. Each consecutive model is more complex. The intraclass correlation is computed with the null model (model1).

```{r tutorial-11, eval=TRUE}                                               
model1 <- fitCyclicMLA(dat=dat3, yvar="intention", id = "subjnr", 
                       ymin = -0.5, ymax = 0.5, step=0.10 )
print(model1)
plot(model1)

ICC <- (as.data.frame(lme4::VarCorr(model1$fit))[1,"vcov"]) / sum(as.data.frame(lme4::VarCorr(model1$fit))[,"vcov"])          
cat("The Intraclass correaltionn (ICC) is: ", ICC)

model2 <- fitCyclicMLA(dat=dat3, random = "intercept",  ncycle = 1,
                       yvar="intention", xvar1="beepnr",xvar2="daynr", id = "subjnr", 
                       ymin = -0.5, ymax = 0.5, step=0.10 )

print(model2)
plot(model2)


model3 <- fitCyclicMLA(dat=dat3, yvar="intention", xvar1="beepnr",xvar2="daynr", id = "subjnr", 
                      ncycle = 1, random = "first",ymin = -0.5, ymax = 0.5, step=0.10 )

print(model3)
plot(model3)

## Fitting two cyclic process: within day and within week

model4 <- fitCyclicMLA(dat=dat3, yvar="intention", xvar1="beepnr", xvar2="daynr",id = "subjnr", 
                        ncycle = 2, random = "first", ymin = -0.5, ymax = 0.5, step=0.10 )
print(model4)
plot(model4)

model5 <- fitCyclicMLA(dat=dat3, yvar="intention", xvar1="beepnr", xvar2="daynr",id = "subjnr",
                       ncycle = 2, random = "all", ymin = -0.5, ymax = 0.5, step=0.10 )

print(model5)
plot(model5)

model6 <- fitCyclicMLA(dat=dat3, yvar="intention", xvar1="beepnr", xvar2="daynr",id = "subjnr", cov = "stress",
                        ncycle = 2, random = "all", ymin = -0.5, ymax = 0.5, step=0.10 )

print(model6)
plot(model6)

# compare the models

anova(model6$fit, model5$fit, model4$fit,model3$fit, model2$fit, model1$fit)  

# ratio residuals and dependent variable

sum( (dat3$intention - predict(model6$fit)) **2) / sum(dat3$intention ** 2)     

```






